A free/open-source machine translation system for English to Kazakh

### 2015-03-04 engkaz-eamt-2015/engkaz-eamt-2015.tex
Mikel's first ironing
git-svn-id: https://svn.code.sf.net/p/apertium/svn/branches/papers/@59213 72bbbca6-d526-0410-a7d9-f06f51895060

\documentclass[11pt]{article}
\usepackage{eamt15}
\usepackage{times}
\usepackage{latexsym}
\setlength\titlebox{6.5cm}    % Expanding the titlebox
%%% YOUR PACKAGES BELOW THIS LINE %%%
\usepackage{url}

@@

\usepackage{polyglossia}
\setdefaultlanguage[variant=australian]{english}
\usepackage{latexsym}
%\usepackage[small,bf]{caption}
\usepackage[bf]{caption}
\usepackage{xltxtra}
\usepackage{times}
\usepackage{fontspec}
\usepackage{footnote}
\usepackage{natbib}
\defaultfontfeatures{Scale=MatchLowercase,Mapping=tex-text}
\usepackage{listings}
\usepackage{microtype}

@@

\newcommand{\com}[1]{\marginpar{\scriptsize #1}}

\title{A free/open-source machine translation system for English to Kazakh}

@@

\begin{document}
\maketitle

\begin{abstract}
This paper presents the current state of development a shallow-transfer rule-based machine translation (MT) system from English to Kazakh. The main syntactic and morphological differences 
between the two languages are presented, and it is showed how the MT 
system was designed to tackle these challenges. We show an evaluation of system coverage and 
translation quality and outline and future work.
\end{abstract}

@@

We currently build rule-based machine translation system, because for statistical machine translation 
system we need large sentence-aligned parallel corpora. There are no freely available, large parallel 
corpora for the English-Kazakh pair. English-Kazakh statistical MT system was built by Google,\footnote{\url{http://translate.google.com}} 
other work in this language pair was made by building a small parallel corpus and using Moses \citep{assylbekov14}, 
and existing commercial systems for English to Kazakh (Sanasoft\footnote{\url{http://www.sanasoft.kz/c/ru/node/47}} and Trident\footnote{\url{http://www.translate.ua/us/on-line}}) both appear to be rule-based. Rule-based machine translation, in which developers 
write up dictionaries and grammatical rules that are applied by an engine, in some 
unusual case will produce better translation.

This paper describes work in progress of development of machine translation system 
for English-to-Kazakh, which developed by using Apertium free/open-source machine 
translation platform \citep{forcada11}. %(http://www.apertium.org).

@@

\begin{figure*}[htbp]
\begin{center}
 \includegraphics[width=0.8\textwidth]{architecture.pdf}
\end{center}
\caption{The pipeline architecture of the Apertium system.}
\label{fig:modules}
%\vspace{-1em}
\end{figure*}

Apertium \citep{forcada11} is a free/open-source rule-based machine translation (MT) platform that 
has been developed since 2005 at the Universitat d'Alacant. At first, it was initially aimed at 
translating texts between closely related languages, then it was extended to deal with less-related 
languages. This platform has next components: machine translation engine, developer's tools, and linguistic 
data for an increasing number of language pairs and they are licensed under the 
free/open-source GNU General Public License (GPL, versions 2 and 3).

@@

\item[Morphological analyser:] For each surface form the morphological analyser generates one or more 
  lexical forms, which consist of: lemma (dictionary or citation form), lexical category (or part-of-speech), 

@@

\item[Part-of-speech (POS) tagger:] This module chooses one of lexical forms of ambiguous word. As can be 
  seen from the previous example, the morphological analyser could deliver more than one lexical form and 
  choosing wrong one could produce errors in translation. POS tagger is based on a statistical model, which 
  based on hidden Markov models and it has been trained on source language texts, which processes the 
  result of the application of  on constraint-grammar rules \citep{karlsson95}, which are used to discard some 
  analyses using rules based on context. 

@@

After  this module, all words have only one morphological analysis.
\item[Lexical transfer:] It reads each lexical form of source language and delivers a corresponding target language 
  lexical form. This module uses a bilingual dictionary. 
%For the English-Kazakh language pair, each lexical form of 

@@

\item[ Lexical selection:] It uses rules for lexical words, which have many translations, to select one of the translations 
  in the target language according to context. The system is described in \cite{tyers12a}.
\item [Structural transfer:] This module uses pattern matching to identify sequences of lexical forms  (phrases or 
  segments), which need syntactical processing to translate grammatical differences between two languages (handling of 
  number, gender, etc.). It uses  files with rules, which specify  the syntactic transformation such as word reorderings, 

@@

\item[Morphological generator:] Delivers the sequence of target-language lexical forms, produced by the structural transfer, 
  to a corresponding sequence of target language surface forms. The morphological generator executes a finite-state transducer 
  generated by compiling a morphological dictionary for the target language.     
  For Kazakh language a finite-state transducer based on two-level rules \citep{washingtonsalimzyantyers14} is used for generation. 
\item[ Post-generator:] Performs some minor orthographical operations in the target language (for instance, it generates the 
  English form cannot from can and not). This module is generated from file with rules which are very   similar in format to dictionary files. 

@@

Apertium uses XML-based formats for linguistic data, which include bilingual and monolingual dictionaries, structural transfer and
lexical selection rules. % and rules for part-of-speech tagging.

@@

Dictionaries are used in lexical processing and morphological analysis, there are use two types: morphological (monolingual) and bilingual.

\subsubsection{Monolingual dictionaries}

English dictionary uses to determine lexical forms of each surface forms, it contains of list of all lexical units, definition 
of source-language alphabet, grammatical categories, which is used to specify lexical forms, for example, like noun, verb, 
gender, etc., each categories have paradigms, they describes groups of correspondences between parts of surface forms and lexical forms.

For Kazakh side, is used morphological transducers, which are based on the Helsinki Finite State Toolkit \citep{hfst/2011}. It 
uses formalisms lexc for defining lexicons, twol formalism for morphophonological rules. Lexc formalism defines different 
word classes and subclasses. And twol determines phonological transformation like vowel harmony, desonorisation. For example, 
depend on preceding vowel, plural form \emph{-LAr} could become -тар or -тер depending on vowel harmony: kitap+tar, but mektep+ter.

\subsubsection{Bilingual dictionary}

Bilingual dictionary provides correspondences between lexical forms of source and target lanugages. Each entry 
consists of English word on the left side with part of speech and corresponding Kazakh word on the right side. 
Some entries could be ambiguous by part of speech, or just have many translations, which will be chosen by 
lexical-selection rules depending on context. This dictionary currently contains 13,130 stems \citep{sundetova13a}.

@@

\subsection{Disambiguation rules}

This kind of rules solves the problem of part-of-speech (POS) ambiguity by Constraint Grammar (CG, \cite{karlsson95}). The 
output of morphological analyser could be ambiguous, and in this case CG rules choose correct analysis from 
multiple analyses. From English to Kazakh POS ambiguity could be only in English side, but also some choice also
depend on Kazakh grammar.

\subsection{Lexicon selection rules }

Lexical-selection rules choose one of the alternative translations of one word depend on the context. The rule 
formalism is described in \cite{tyers12a}. Alternative translations are defined in bilingual dictionary by 
multiple entries of one word. For exmaple, adjective ``beautiful'' could be translated as ``сұлу'', if 
following noun is subject: ``beautiful girl'' -- ``сұлу қыз''; or ``әдемі'', ``көркем'', if the noun means 
place or: ``beautiful mountains'' -- ``әдемі таулар''.  Table~\ref{table:lexsel} demonstrates some 
lexical-selection rules:

\begin{table*}
 \centering
 \begin{tabular}{|l|l|l|l|}
    \hline
    \textbf{Word} & \textbf{Translation} & \textbf{Condition} & \textbf{Example} \\
    \hline
    \multirow{4}{*}{residence} & \multirow{2}{*}{мекен}      & \multirow{2}{*}{--} & I live in a residence. \\
                               &                             &                     & (Мен) мекенде өмір сүремін. \\
                               & \multirow{2}{*}{резиденция} & \multirow{2}{*}{\_ of \_ president} & I see the residence of the president \\ 
                               &                             &                                     & (Мен) президенттің резиденциясын көремін \\
    \hline
    \multirow{4}{*}{boot} & \multirow{2}{*}{бәтеңке} & -- & I bought boots. \\
                          &                          &     &  Мен бәтеңкелерді сатып алдым \\

                          & \multirow{2}{*}{жүксалғыш} & \_ of \_ car &  He opened the boot of the car  \\ 
                          &           &              & Ол машинаның жүксалғышын ашты. \\
    \hline
    \multirow{4}{*}{anything} & \multirow{2}{*}{бір нәрсе} & -- & I see anything \\ 
                              &           &    & Мен бір нәрсе көремін \\
                              & \multirow{2}{*}{еш нәрсе} & not \_[verb] & I do not do anything \\ 
                              &          &             &  Мен еш нәрсе істемеймін. \\

    \hline
 \end{tabular}
  \caption{Example lexical-selection rules}
  \label{table:lexsel}
\end{table*}

\subsection{Transfer rules}

For English–Kazakh system this stage uses three stages to improve the granularity of structural-transfer 
rules \citep{sundetova13b}:

@@

This section describes the current structural transfer in \texttt{apertium-eng-kaz} (revision ..., ). English to Kazakh 
chunker rules, interchunk rules and additional clean-up stage will be described in detail next 3 sections.

\subsubsection{Chunker}

In the first round of structural transfer rules do some local operations to cut sentences into chunks, like short noun 
phrases, adjective phrases, verb phrases and adpositional phrases (that is, prepositional phrases in English and 
postpositional phrases in Kazakh).
Chunking rules, of which there are currently 168, identify 8 kinds of chunks and translate them into Kazakh as much 
as possible, leaving some minor operations to be performed in later stages of structural transfer (for instance, the 
case of noun phrases). Table~\ref{table:nps} describes what kind of noun phrases could appear in sentences and how it could be detected.

@@

\begin{savenotes}
\begin{table*}
  \centering
  \begin{tabular}{|l|l|l|l|}
    \hline
    \textbf{Phrase type} & \textbf{Description} & \textbf{Examples} & \textbf{Chunk tags} \\
    \hline
    \multirow{5}{*}{\texttt{NP}} & \multirow{5}{*}{Noun phrase}    & noun & number*, person, \\
                                 &                                 & determiner-noun & possessor*, case\\
                                 &                                 & numeral-noun & \\
                                 &                                 & adjective-noun & \\
                                 &                                 & determiner-adjective-noun & \\
    \hline
    \multirow{5}{*}{\texttt{VP}} & \multirow{5}{*}{Verb phrase}    & finite-verb & number, person, \\
                                 &                                 & have-participle & tense/mood, possessive*, \\
                                 &                                 & be-gerund & negation* \\
                                 &                                 & must-infinitive & \\ 
                                 &                                 & want\_to-infinitive & \\ 
    \hline
    \multirow{5}{*}{\texttt{PP}} & \multirow{5}{*}{Postpositional phrase} & preposition-noun & number, person, \\
                                 &                                        & preposition-determiner-noun & possessor*, case \\
                                 &                                        & preposition-numeral-noun & \\
                                 &                                        & preposition-adjective-noun & \\
                                 &                                        & preposition-determiner-adjective-noun & \\
    \hline
    \multirow{2}{*}{\texttt{GenP}} & \multirow{2}{*}{Genitive phrase}  & As with \texttt{PP} except & number, person, \\
                                   &                                   & with the preposition `of'. & possessor*, case \\
    \hline
    \multirow{2}{*}{\texttt{AdjP}} & \multirow{2}{*}{Adjective phrase}  & adjective & n/a \\
                                   &                                    & ``more''--adjective & \\
    \hline
    \multirow{2}{*}{\texttt{SupP}} & \multirow{2}{*}{Superlative phrase} & adjective--``est'' & possessor*,\footnote{Treated as a noun phrase with an implied noun (``the largest [book]'')} case \\
                                   &                                     & the--``most''--adjective & \\ 
    \hline
%    \texttt{Q\_mark} & Question mark  & & \\
%    \hline
  \end{tabular}
  \caption{Phrase types in the English-Kazakh chunker}
  \label{table:phrases} 
\end{table*}
\end{savenotes}

As it is shown above, the most probable sequence of world classes are built phrases and 
for English-Kazakh chunker there are used next kind of phrases:

@@

  possessor (to be determined, as the noun кітап ('book') could receive a 3rd-person possesive 
  ending (кітабы) later if the context were, for instance, the large book of animals, аңдардың үлкен кітабы), 
  and case (to be determined as it could be, for instance, accusative in I saw the large book, Мен үлкен кітапты көрдім ). Also 
  noun form of the verb, for instance,  gerund form is determined as NP phrase (Her playing is good – Оның ойыны жақсы).

@@

  there are three possible outcomes with different cases:

@@

    \item Translating the locative ``-{D}{A}'',\footnote{{D} can be д or т, and {A} can be е or а, depending on the 
        phonological context.} ablative ``-{D}{A}н'', etc., except the genitive ``-{N}{I}ң'' cases: 
        [PP [P from ] [NP the five cars] ] $\rightarrow$ [PP [NP бес көлік] [P -тен ] ]
    \item With nouns such as аст `under', үст `on', etc. which takes possessive from main noun:  
        [PP [P under ] [NP the garden] ] $\rightarrow$ [PP [NP [GenP [NP бақша ][P -ның]] [NP астын]] [P -да ] ]

@@

\item[Verb phrases:] Translation verb phrases into Kazakh  are not always be straightforward, for instance: 
  present simple and future are rendered using the same tense in Kazakh (I play $\rightarrow$ Мен ойнаймын;  I 
  will play $\rightarrow$ Мен ойнаймын), this ambiguity problem may be solved by lexical selection, 
  which will choose right tense by watching context; tenses expressing continued activity, such as the 

@@

  modal verbs are translated by adding adjectives, which means ``necessary'' (керек) or ``proper'' (жөн), 

@@

\item The adjective alone, marked AdjP: [AdjP big ] $\rightarrow$ [AdjP үлкен] 

@@

   chooses the comparative suffix ``-{I}р{A}{K}'': [AdjP  more  interesting ]  $\rightarrow$ [AdjP қызығырақ]

@@

   is built using ``ең'' + adjective:  [SupP  the biggest ]  $\rightarrow$ [SupP ең үлкен] 
   [SupP  the most important]   $\rightarrow$ [SupP ең маңызды]

@@

modified by a genitive phrase: the most beautiful of people $\rightarrow$ адамдардың ең әдемісі); one could 
say that they are treated as NPs with an implied noun.

@@

\subsubsection{Interchunk processing}

The second round of structural transfer is currently performed by a proof-of-concept set of 90 rules, representative of following operations:
\begin{itemize}
\item Inter-chunk agreement between subject noun phrase and verb phrase in their person and number.
\item Assigning case to noun phrases (which are generated without case by the chunker): for instance, 
  accusative case for objects (I bought the table $\rightarrow$ Мен үстелді  сатып алдым), genitive case for 
  obligatory constructs (I must see $\rightarrow$ менің көруім керек),  dative case for the verb to 
  need (I need a doctor $\rightarrow$Маған дәрігер керек), locative case for 
  possession (I have a book $\rightarrow$ Менде кітап бар), etc.
\item Reordering: placing of object before verb ( I[1] bought[2] the table[3] $\rightarrow$ Мен[1] үстелді[3] сатып алдым[2]), 
  placing of prepositional pharses before the verb  (They[1] played[2] on top of the tree [3] $\rightarrow$ Олар[1] ағаштың үстінде[3] ойнады [2]), etc.
\item Adding question word ма/ме/ба/бе in the end of question, if question mark chunk ``?'' is detected (Did<VP\_ques> you<NP> watch<VP> last film<NP> ?<Q\_m> $\rightarrow$ Сіз<NP> соңғы фильмді<NP> көрдіңіз<VP> бе<ques> ?<Q\_m>).
\item Place possessive for long noun + noun + noun structures(The university of city of Kazakhstan -- Қазақстан қаласының университеті).
\end{itemize}

@@

\subsubsection{Postchunk}

Additional cleaning stage was created for removing extra tags which did not determine through the chunker and 
interchunk steps, for instance, noun was not acting as object and did not get accusative case, so not determined 
tag was \texttt{<CD>} after ``cleaning'' it will be set on nominative case; at this step, also carried out 
agreement between noun or adjective and copula ``е''(Мен дәрігер+е<p1 person singular>), vowel and consonants 
agreement between last word of question and question particle (Сіз келдіңіз + ме? - Сіз келдіңіз + бе?).

@@

Evaluating the output of the system was measured with BLEU \citep{papineni02} and Word Error Rate 
metric \citep{levenshtein/1966}. We chose a short text in English, and by postediting output 
of English-Kazakh system, was build parallel text. The output of each of the machine
translation systems was posteditted independently to avoid biasing one particular system.

As can be seen,  coverage of dictionary is good, average score is 94\%. Our system outperforms the other 
available RBMT system, but fall short of the state-of-the-art performance. However, unlike the state of 
the art, which uses unavailable corpora, our system is free and open and can be improved.

>>>>

\documentclass[11pt]{article}
\usepackage{eamt15}
%\usepackage{times}
\usepackage{latexsym}
\setlength\titlebox{6.5cm}    % Expanding the titlebox
%%% YOUR PACKAGES BELOW THIS LINE %%%
\usepackage{url}

@@

\usepackage{polyglossia}
\usepackage{todonotes}
\setdefaultlanguage[variant=australian]{english}
\usepackage{latexsym}
%\usepackage[small,bf]{caption}
\usepackage[bf]{caption}
\usepackage{xltxtra}
\usepackage{times}
\usepackage{fontspec}
\usepackage{footnote}
\usepackage{natbib}
\defaultfontfeatures{Scale=MatchLowercase,Mapping=tex-text}
\usepackage{listings}
\usepackage{microtype}

@@

\newcommand{\com}[1]{\marginpar{\scriptsize #1}} 
    \setcounter{secnumdepth}{4}
\title{A free/open-source machine translation system for English to Kazakh}

@@

\begin{document}
\maketitle 
\renewcommand{\baselinestretch}{0.97} % the undetectable 3% squeeze

\begin{abstract}\todo{Decide authors, and put them in phantom boxes}
This paper presents the current state of development a shallow-transfer rule-based machine translation (MT) system from English to Kazakh. The main syntactic and morphological differences 
between the two languages are presented, and it is showed how the MT 
system was designed to tackle these challenges. We show an evaluation of system coverage and 
translation quality and outline and future work.
\end{abstract}

@@

We set out to build a rule-based machine translation system, because for statistical machine translation 
system we need large sentence-aligned parallel corpora, which are not freely available
for the English--Kazakh pair. An English--Kazakh statistical MT system has however recently been built by Google,\footnote{\url{http://translate.google.com}}; 
other work in this language pair was performed by building a small parallel corpus and using Moses \citep{assylbekov14}. Existing commercial systems for English to Kazakh (Sanasoft\footnote{\url{http://www.sanasoft.kz/c/ru/node/47}} and Trident\footnote{\url{http://www.translate.ua/us/on-line}}) both appear to be rule-based. 
Rule-based machine translation, in which developers 
write up dictionaries and grammatical rules that are applied by an engine, in some 
unusual case will produce better translation.
\todo{What does the last sentence mean? Clarify!}
This paper describes work in progress of development of machine translation system 
for English--Kazakh using the Apertium free/open-source machine 
translation platform \citep{forcada11}. %(http://www.apertium.org).

@@

\begin{figure*}[htbp]
\begin{center}
 \includegraphics[width=0.8\textwidth]{architecture.pdf}
\end{center}
\caption{The pipeline architecture of the Apertium system.}
\label{fig:modules}
\vspace{-1em}
\end{figure*}

Apertium \citep{forcada11} is a free/open-source rule-based machine translation (MT) platform that 
has been developed since 2005, started at the Universitat d'Alacant. At first, it was initially aimed at 
translating texts between closely related languages, then it was extended to deal with less-related 
languages. This platform has the following components: machine translation engine, developer's tools, and linguistic 
data for an increasing number of language pairs and they are licensed under the 
free/open-source GNU General Public License (GPL, versions 2 and 3).
\todo{If space allowed, it would be nice to add a running example as in the Apertium paper}
\begin{description}\itemsep 0ex

@@

\item[Morphological analyser:] For each surface form  (SF) the morphological analyser generates one or more 
  lexical forms (LF), which consist of: lemma (dictionary or citation form), lexical category (or part-of-speech), 

@@

\item[Part-of-speech (POS) tagger:] This module chooses one of the LF of an ambiguous SF. 
%As can be 
%  seen from the previous example, the morphological analyser could deliver more than one LF and 
%  choosing wrong one could produce errors in translation. 
\todo{I have removed some text and  the statistical PoS tagging stuff, as it is not used}
%POS tagger is based on a statistical model, which 
%  based on hidden Markov models and it has been trained on source language texts, which processes the 
%  result of the application of  on constraint-grammar rules \citep{karlsson95}, which are used to discard some 
%  analyses using rules based on context. 

@@

%After  this module, all words have only one morphological analysis.
\item[Lexical transfer:] It reads each source-language LF and delivers the corresponding target-language 
  LFs. This module uses a bilingual dictionary. 
%For the English--Kazakh language pair, each lexical form of 

@@

\item[ Lexical selection:] It uses rules to select one of the target-language LFs, as described in \cite{tyers12a}.
\item [Structural transfer:] This module uses pattern matching to identify sequences of LF (phrases or 
  segments), which need syntactical processing 
%to deal 
%with grammatical differences between two languages (handling of 
%  number, gender, etc.). 
It uses files with rules, which specify  the syntactic transformation such as word reorderings, 

@@

\item[Morphological generator:] Transforms the sequence of target-language LFs forms, produced by the structural transfer, 
  to a corresponding sequence of target-language SFs. 
% The morphological generator executes a finite-state transducer 
For Kazakh, a finite-state transducer based on two-level rules \citep{washingtonsalimzyantyers14} is used for generation. 
\item[Post-generator:] Performs some minor orthographical operations in the target language. This module is not used in the English--Kazakh system.\todo{Or is it?}

@@

Apertium mainly uses XML-based formats for linguistic data, which include bilingual and monolingual dictionaries, structural transfer and
lexical selection rules.

@@

Dictionaries are used in lexical processing: monolingual dictionaries
for morphological analysis of English and generation of Kazakh and
bilingual dictionaries for English--Kazakh lexical transfer.

The English dictionary is used to determine lexical forms of each surface
forms, it contains of list of all lexical units, definition of
source-language alphabet, grammatical categories, which is used to
specify lexical forms, for example, like noun, verb, gender, etc.;
each category has paradigms describing groups of correspondences
between parts of SFs and LFs.

The Kazakh dictionary is turned into a
morphological transducer, using the Helsinki Finite
State Toolkit \citep{hfst/2011}. It uses the \texttt{lexc} formalism
for defining lexicons through word classes and subclasses, and the
\texttt{twol} formalism for morphophonological rules such as vowel
harmony, desonorization, nasalization, etc.  Depending on the
preceding vowels and consonants, the plural suffix \emph{-LAr}\footnote{Uppercase Latin letters are used for archiphonemes (actually archigraphemes) that are realised as phonemes (actually graphemes) after morphophonological rules have been applied} could
become \emph{-тар}, \emph{-тер}, \emph{-lar}, \emph{-ler}, \emph{-dar}, \emph{-ler} depending on vowel harmony: kitap+tar,
mektep+ter, etc. \todo{Aida, please typeset this example in Cyrillic; braces are not used here, but are used later; homogeneize notation in paper.}

%\subsubsection{Bilingual dictionary}

The English--Kazakh bilingual dictionary provides correspondences between English LFs and Kazakh LFs. Ambiguity is allowed: one of the LFs will be chosen by 
lexical-selection rules depending on context. This dictionary currently contains 13,130 stems \citep{sundetova13a}.\todo{later you say 13,135}

@@

\subsubsection{Disambiguation rules}

This kind of rules solves the problem of part-of-speech (PoS) ambiguity using the  Constraint Grammar (CG, \cite{karlsson95}) formalism. CG rules choose one of the LFs obtained for each SF. %From English to Kazakh POS ambiguity could be only in English side, but also some choice also
%depend on Kazakh grammar.\todo{I removed an unclear sentence here}

\subsubsection{Lexical selection rules }

Lexical-selection rules choose one of the alternative target-language LFs corresponding to one source-language LF, as described in \cite{tyers12a}. Alternative translations are defined in bilingual dictionary by 
multiple entries for each source-language LF. For example, the adjective \emph{beautiful} could be translated as \emph{сұлу}, if 
the following noun is a person: \emph{beautiful girl} \(\to\) \emph{сұлу қыз}; or as \emph{әдемі} or  \emph{көркем}, if the following noun means 
place or \todo{or what?} \emph{beautiful mountains} \(\to\) \emph{әдемі таулар}.  Table~\ref{table:lexsel} shows some 
lexical-selection rules:\todo{but this rule is not in the table!!!}
\begin{table*}
 \centering
 \begin{tabular}{|l|l|l|l|}
    \hline
    \textbf{SL word \(w\)} & \textbf{TL word} & \textbf{Context} & \textbf{Example} \\
    \hline
    \multirow{4}{*}{residence} & \multirow{2}{*}{мекен}      & \multirow{2}{*}{default} & I live in a residence. \\
                               &                             &                     & (Мен) мекенде өмір сүремін. \\
                               & \multirow{2}{*}{резиденция} & \multirow{2}{*}{\(w\)\_ of \_ president} & I see the residence of the president \\ 
                               &                             &                                     & (Мен) президенттің резиденциясын көремін \\
    \hline
    \multirow{4}{*}{boot} & \multirow{2}{*}{бәтеңке} & default & I bought boots. \\
                          &                          &     &  Мен бәтеңкелерді сатып алдым \\

                          & \multirow{2}{*}{жүксалғыш} & \(w\) \_ of \_ car &  He opened the boot of the car  \\ 
                          &           &              & Ол машинаның жүксалғышын ашты. \\
    \hline
    \multirow{4}{*}{anything} & \multirow{2}{*}{бір нәрсе} & default & I see anything  
\\ 
                              &           &    & Мен бір нәрсе көремін \\
                              & \multirow{2}{*}{еш нәрсе} & not \_[verb] & I do not do anything \\ 
                              &          &             &  Мен еш нәрсе істемеймін. \\

    \hline
 \end{tabular}
  \caption{Example lexical-selection rules}
  \label{table:lexsel}
\end{table*}
\todo{I do not understand the \emph{I see anything} example. Does it mean \emph{anything} as \emph{whatever}? I do not know how to put the \(w\) in this example}@

\subsubsection{Transfer rules}

For English--Kazakh transfer is performed in three stages \citep{sundetova13b}:

@@

This section describes the current structural transfer in \texttt{apertium-eng-kaz} (revision ..., ).\todo{Complete! 58998? Move to results?} English--Kazakh 
chunker rules, interchunk rules and an additional clean-up stage will be described in detail in the following 3 sections.

\paragraph{Chunker:}\todo{This should have 4 numerals, but I don't know how to do this}
In the first round of structural transfer, rules segment sentences into chunks, like short noun 
phrases, adjective phrases, verb phrases and adpositional phrases (that is, prepositional phrases in English and 
postpositional phrases in Kazakh).
Chunking rules, of which there are currently 168, identify 8 kinds of chunks and translate them into equivalent Kazakh chunks, leaving some adaptations to be performed in later stages of structural transfer (for instance, the 
morphological case of noun phrases). Table~\ref{table:nps} describes what kind of noun phrases could appear in sentences and how it could be detected.\todo{This table does not add much. the first and second column of the table are essentially the same: it would be more interesting to give the Kazakh sequence of PoS tags, perhaps with an example}

@@

\begin{savenotes}
\begin{table*}
  \centering
  \begin{tabular}{|l|l|l|l|}
    \hline
    \textbf{Phrase type} & \textbf{Description} & \textbf{Examples} & \textbf{Chunk tags} \\
    \hline
    \multirow{5}{*}{\texttt{NP}} & \multirow{5}{*}{Noun phrase}    & noun & number*, person, \\
                                 &                                 & determiner-noun & possessor*, case\\
                                 &                                 & numeral-noun & \\
                                 &                                 & adjective-noun & \\
                                 &                                 & determiner-adjective-noun & \\
    \hline
    \multirow{5}{*}{\texttt{VP}} & \multirow{5}{*}{Verb phrase}    & finite-verb & number, person, \\
                                 &                                 & have-participle & tense/mood, possessive*, \\
                                 &                                 & be-gerund & negation* \\
                                 &                                 & must-infinitive & \\ 
                                 &                                 & want\_to-infinitive & \\ 
    \hline
    \multirow{5}{*}{\texttt{PP}} & \multirow{5}{*}{Postpositional phrase} & preposition-noun & number, person, \\
                                 &                                        & preposition-determiner-noun & possessor*, case \\
                                 &                                        & preposition-numeral-noun & \\
                                 &                                        & preposition-adjective-noun & \\
                                 &                                        & preposition-determiner-adjective-noun & \\
    \hline
    \multirow{2}{*}{\texttt{GenP}} & \multirow{2}{*}{Genitive phrase}  & As with \texttt{PP} except & number, person, \\
                                   &                                   & with the preposition `of'. & possessor*, case \\
    \hline
    \multirow{2}{*}{\texttt{AdjP}} & \multirow{2}{*}{Adjective phrase}  & adjective & n/a \\
                                   &                                    & ``more''--adjective & \\
    \hline
    \multirow{2}{*}{\texttt{SupP}} & \multirow{2}{*}{Superlative phrase} & adjective--``est'' & possessor*, case \\
                                   &                                     & the--``most''--adjective & \\ 
    \hline
%    \texttt{Q\_mark} & Question mark  & & \\
%    \hline
  \end{tabular}
  \caption{Phrase types in the English-Kazakh chunker}
  \label{table:phrases} 
\end{table*}
\end{savenotes}
As it is shown above, the most probable sequence of world classes are built phrases and 
for English-Kazakh chunker there are used next kind of phrases:

@@

  possessor (to be determined, as the noun \emph{кітап} ('book') could receive a 3rd-person possesive 
  ending (\emph{кітабы}) later if the context were, for instance, \emph{the large book of animals}, \emph{аңдардың үлкен кітабы}), 
  and case (to be determined as it could be, for instance, accusative in \emph{I saw the large book}, \emph{Мен үлкен кітапты көрдім} ). Also, the
  noun form of the verb, for instance,  gerund form is determined as NP phrase (\emph{Her playing is good} – \emph{Оның ойыны жақсы}).\todo{Make sure all English and Kazakh examples are in italics}

@@

  there are three possible outcomes with different cases:\todo{remove braces for consistency with the previous notation for archiphonemes: avoid repeating the explanation in the footnote.}

@@

    \item Translating into any of the morphological cases: locative ``-{D}{A}'',\footnote{{D} can be д or т, and {A} can be е or а, depending on the 
        phonological context.} ablative ``-{D}{A}н'', etc., except the genitive ``-{N}{I}ң'' case: 
        [PP [P \emph{from} ] [NP \emph{the five cars}] ] $\rightarrow$ [PP [NP \emph{бес көлік}] [P \emph{-тен} ] ]
    \item Using postpositional constructs based on positional nouns such as \emph{аст} (`under'), \emph{үст} (`on'), etc., which take the possessive from the main noun:  
        [PP [P \emph{under} ] [NP \emph{the garden}] ] $\rightarrow$ [PP [NP [GenP [NP \emph{бақша} ][P \emph{-ның}]] [NP \emph{аст\textbf{ын}}]] [P \emph{-да} ] ]

@@

\item[Verb phrases:] Translation of English verb phrases into Kazakh  are not always be straightforward, for instance: 
  % present simple and future are rendered using the same tense in Kazakh (I play $\rightarrow$ Мен ойнаймын;  I 
  % will play $\rightarrow$ Мен ойнаймын), this ambiguity problem may be solved by lexical selection, 
  % which will choose right tense by watching context; 
\todo{I removed text that pertained to kaz--eng}
tenses expressing continued activity, such as the 

@@

  modal verbs are translated by adding adjectives, which means ``necessary'' (керек) or ``proper'' (жөн), \todo{All examples in italics. This part is hard to read. The choice of auxilliary is not justified.}

@@

\item The adjective alone, marked AdjP: [AdjP \emph{big} ] $\rightarrow$ [AdjP \emph{үлкен}] 

@@

   chooses the comparative suffix ``-{I}р{A}{K}'': [AdjP  \emph{more  interesting} ]  $\rightarrow$ [AdjP \emph{қызығ\emph{ырақ}}]

@@

   is built using \emph{ең} + adjective:  [SupP  the \emph{biggest} ]  $\rightarrow$ [SupP \emph{ең үлкен}] 
   [SupP  \emph{the most important}]   $\rightarrow$ [SupP \emph{ең маңызды}]

@@

modified by a genitive phrase: \emph{the most beautiful of people} $\rightarrow$ \emph{адамдардың ең әдемі\textbf{сі}}); they are treated as a noun phrase with an implied noun (``the largest [book]'').
\todo{The table of phrase types is not referred to in the text; also, the meaning of ``*'' is not explained}

@@

\paragraph{Interchunk processing:}
\todo{Table~\ref{table:vps} should be typeset with examples in italics and making sure that the margins are respected}

The second round of structural transfer is currently performed by a %proof-of-concept 
set of 90 rules, representative of following operations:\todo{I have removed \emph{proof-of-concept} here}
\begin{itemize}
\item Inter-chunk agreement between subject noun phrase and verb phrase in their person and number.
\item Assigning case to noun phrases (which are generated without case by the chunker): for instance, 
  accusative case for objects (\emph{I bought the table} $\rightarrow$ \emph{Мен үстелді  сатып алдым}), genitive case for 
  obligatory constructs (\emph{I must see} $\rightarrow$ \emph{менің көруім керек}),  dative case for the verb \emph{to 
  need} (\emph{I need a doctor} $\rightarrow$ \emph{Маған дәрігер керек}), locative case for 
  possession (\emph{I have a book} $\rightarrow$ \emph{Менде кітап бар}), etc.\ 
\item Reordering: placing of object before verb ( I[1] bought[2] the table[3] $\rightarrow$ Мен[1] үстелді[3] сатып алдым[2]), \todo{italics for examples please!}
  placing of prepositional pharses before the verb  (They[1] played[2] on top of the tree [3] $\rightarrow$ Олар[1] ағаштың үстінде[3] ойнады [2]), etc.\ 
\item Adding question word ма/ме/ба/бе in the end of question, if the question mark chunk ``?'' is detected (Did<VP\_ques> you<NP> watch<VP> last film<NP> ?<Q\_m> $\rightarrow$ Сіз<NP> соңғы фильмді<NP> көрдіңіз<VP> бе<ques> ?<Q\_m>).\todo{Notation: use brackets as above!}
\item Place possessive for long noun + noun + noun structures(The university of city of Kazakhstan -- Қазақстан қаласының университеті).\todo{Explain better, as this is a rule to avoid repeating -NIng, -NIng, etc. All possessives are in place}
\end{itemize}

@@

\paragraph{Postchunk:}\todo{Nothing is said about postchunk --- this talks about the additional cleaning stage}
An additional cleaning stage was created for removing extra tags which were ot determined during the chunker and 
interchunk steps, for instance, a noun was not determined to be acting as an object and therefore did not get accusative case; such a noun would have been assigned the  \texttt{<CD>} (``case to be determined'') tag, which after ``cleaning'' would be set to nominative. Other operations  carried out at this stage ensure
the agreement between noun or adjective and the copula verb ``е''(\emph{Мен дәрігер+е}<p1 person singular>), \todo{add result for copula} or deciding the actual form of the question particle according to the preceding word (\emph{Сіз келдіңіз} + \emph{ме}? \(\to\) \emph{Сіз келдіңіз} + \emph{бе}?).

@@

The output of the system was evaluated with BLEU \citep{papineni02} and the word error rate 
metric \citep{levenshtein/1966}. We chose a short text in English, and by postediting output 
of English--Kazakh system, a build parallel text was built. The output of each of the machine
translation systems was postedited independently to avoid biasing one particular system.

As can be seen,\todo{where!} the coverage of dictionary is good, average score is 94\%. Our system outperforms the other 
available RBMT system, but fall short of the state-of-the-art performance.\todo{represented by...} However, unlike the state of 
the art, which uses unavailable corpora, our system is free and open and can be improved.

MT systems have some mistakes in translations, which was shown in <...> brackets.\todo{you used *!} Google MT system 
has a higher BLEU score, but in translation separate phrases, it has common error: it does not assign 
the right possessive and case. Sanasoft system has more errors with choosing not right translations 
and out of vocabulary words.
